The Weakly Nonlinear Magnetorotational Instability

### 2016-08-30 paper/widegap_mri_paper.tex
new parameter space figure in paper

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_paramspace_crit_params_Pm_1E-3}
\caption{Growth rates in the ($\reym, k_z$) plane. Color map shows growth rate found by solving the linear eigenvalue problem for each ($\reym, k_z$) in the grid. The eigenvalue problem was solved for the widegap parameters listed in Table \ref{table:parameters}. Overlaid contours show growth rates at [-2E-3, -1E-3, -1E-5, 1E-5, 1E-3, 2E-3], where dashed contours represent negative values. The gray dotted line shows the interpolated marginal stability curve. The critical parameters $\reym_c = 0.84$ and $k_c = 0.0127$ correspond to the smallest parameter values that yield a zero growth rate.}\label{fig:growth_rates}
\end{figure}

@@

and define the magnetic vector potential $A$ analogously. These definitions automatically satisfy Equations \ref{eq:incompressibility} and \ref{eq:solenoid} for axisymmetric disturbances. We note that in the linearized equations, streamfunctions of the form $u_x = \partial_z \Psi$, $u_z = -(\partial_r + \frac{1}{r}) \Psi$, and the corresponding definitions of the magnetic vector potential, are convenient choices, but this does not hold true for the nonlinear terms because of the incommutability of $\partial_r$ and $\partial_r + \frac{1}{r}$.

@@

Thus through judicious choice of cylinder rotation rates, we can set $q(r_0) = 3/2$, for quasi-Keplerian flow. Note that the narrow gap approximation imposes a linear shear (constant $q$). Our base velocity is $\mathbf{u_0} = r\Omega(r) \phihat$. We initialize a magnetic field $\mathbf{B_0} = B_0 \zhat + B_0 \xi \frac{r_0}{r} \phihat$, so that the base magnetic field is axial when $\xi = 0$ and otherwise helical.

In this work we will focus our findings on two fiducial parameter sets, one for the standard MRI (hereafter SMRI) where $\xi = 0$ and one for the helical MRI (hereafter HMRI). We choose the SMRI parameters to be comparable to the case explored in \citei{Umurhan:2007hs}, with the geometric parameters of \citei{Goodman:2002ix}. The HMRI parameters were chosen to be comparable to \citei{Hollerbach:2005tr}. Our fiducial parameters are described in Table \ref{table:parameters}.

\begin{widetext}
\begin{table*}[ht]
\normalsize
\caption{Fiducial parameters for MRI runs} \label{table:parameters}
\centering
%\resizebox{\textwidth}{!}
{\begin{tabular}{rlrrrrrrr}
  \hline
 & $\xi$ & $Pm$ & $\beta$ & $\Omega_2/\Omega_1$ & $R_1/R_2$ & radial magnetic b.c. \\ 
  \hline\hline
%Standard MRI & 0 & 1E-3 & 25 & 0.18 & 0.33 & conducting \\ 
Standard MRI & 0 & 1.6E-6 & 41.2 & 0.121 & 0.33 & conducting \\ 
Helical MRI & 4 & 1E-6 & 1.7E-2 & 0.27 & 0.5 & insulating\\ 
   \hline
\end{tabular}}
\end{table*}

Our perturbed system is

@@

and $B = 0$ at $r = r_1, r_2$ (see \citei{Willis:2002bh}). Here, $I_n$, $K_n$ are the modified Bessel functions of the first and second kind, respectively.

@@

Just as in the weakly nonlinear analyses of \citei{Umurhan:2007hs} and \citei{Clark:2016}, we tune the system away from marginality by taking $B_0 \rightarrow B_0\left(1 - \epsilon^2\right)$, where $\epsilon \ll 1$. We parameterize scale separation as $Z = \epsilon z$ and $T = \epsilon^2 t$, where $Z$ and $T$ are slowly varying spatial and temporal scales, respectively. We group the fluid variables into a state vector $\mathbf{V} = \left[\Psi, u, A, B\right]^{\mathrm{T}}$, such that the full nonlinear system can be expressed as

@@

where $\mathcal{D}$, $\mathcal{L}$, and $\widetilde{\mathcal{G}}$ are matrices defined in Appendix \ref{app:basic_equations}, and $\mathbf{N}$ is a vector containing all nonlinear terms defined in Appendix \ref{app:nonlinear_terms}. We then expand the variables in a perturbation series $\mathbf{V} = \epsilon \mathbf{V}_1 + \epsilon^2 \mathbf{V}_2 + \epsilon^3 \mathbf{V}_3 + h.o.t.$ The perturbed system can then be expressed at each order by the equations

@@

See Appendix \ref{app:basic_equations} for the definition of matrices and nonlinear vectors, and a thorough derivation. We emphasize that Equations \ref{eq:ordere} - \ref{eq:ordere3} have the same form as these equations in the narrow gap case, although the matrices (which contain all radial derivatives) are significantly different in this wide gap formulation. This is because we do not have slow variation in the radial dimension. The slow variation in $Z$ and $T$ are parameterized as an amplitude function $\alpha(Z, T)$ which modulates the flow in these dimensions. This parameterization coupled with the boundary conditions lead us to an ansatz linear solution $\mathbf{V}_1 = \alpha(Z, T) \mathbb{V}_{11}(r) e^{i k_z z} + c.c.$, where the radial variation is contained in $\mathbb{V}_{11}$.

We solve the equations at each order using Dedalus, an open source pseudospectral code (Burns et al. in prep). Because our domain is bounded only radially and not vertically, we solve the radial dimension on a basis of Chebyshev polynomials, and the vertical dimension on a Fourier series basis. We solve Equation \ref{eq:ordere} as a linear eigenvalue problem, and Equations \ref{eq:ordere2} and \ref{eq:ordere3} as linear boundary value problems. %should i mention spurious eigenvals? 
To analyze the MRI system at marginality, we fix the parameters listed in Table \ref{table:parameters} and determine the critical $\reym$ and $k_z$ by repeatedly solving Equation \ref{eq:ordere} to determine the smallest parameter values for which the fastest growing mode is zero. That is, we solve the linear eigenvalue problem for eigenvalues $\sigma = \gamma + i \omega$. Figure \ref{fig:growth_rates} shows linear MRI growth rates $\gamma$ in the $(\reym, k_z)$ plane. For the fiducial standard MRI parameters in Table \ref{table:parameters} we find critical parameters $\reym_c = 0.84$, $k_c = 0.0127$.

@@

We similarly define appropriate boundary conditions for the insulating case. We take the inner product of the adjoint homogenous solution with Equation \ref{eq:ordere3} to find

\beq
\mathcal{L}\mathbf{V}_3 + \xi \widetilde{\mathcal{H}} \mathbf{V}_3 + \mathcal{D} \partial_T \mathbf{V}_1 + \widetilde{\mathcal{L}}_1 \partial_Z \mathbf{V}_2\nonumber 
+ \xi \mathcal{H}\partial_Z \mathbf{V}_2 + \widetilde{\mathcal{L}}_2 \partial_Z^2 \mathbf{V}_1 - \xi \widetilde{\mathcal{H}} \mathbf{V}_1 + \widetilde{\mathcal{G}} \mathbf{V}_1 + \mathbf{N}_3 = 0.
\eeq

>>>>

%\begin{widetext}
\begin{table*}[ht]
\normalsize
\caption{Fiducial parameters for MRI runs} \label{table:parameters}
\centering
\begin{tabular}{rlrrrrrrr}
  \hline
 & $\xi$ & $Pm$ & $\beta$ & $\Omega_2/\Omega_1$ & $R_1/R_2$ & radial magnetic b.c. \\ 
  \hline\hline
Standard MRI & 0 & 1.6E-6 & 41.2 & 0.121 & 0.33 & conducting \\ 
Helical MRI & 4 & 1E-6 & 1.7E-2 & 0.27 & 0.5 & insulating\\ 
   \hline
\end{tabular}
\end{table*} 
%\end{widetext}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_paramspace_crit_params_Pm_16E-7}
\caption{Growth rates in the ($\reym, k_z$) plane. Color map shows growth rate found by solving the linear eigenvalue problem for each ($\reym, k_z$) in the grid. The eigenvalue problem was solved for the widegap parameters listed in Table \ref{table:parameters}. Overlaid contours show growth rates at [ -8E-4, -1.3E-4, 1.3E-4, 8E-4, 1.5E-3], where dashed contours represent negative values. The gray dotted line shows the interpolated marginal stability curve. The critical parameters $\reym_c = 3.31$ and $k_c = 0.902$ correspond to the smallest parameter values that yield a zero growth rate.}\label{fig:growth_rates}
\end{figure}

@@

and define the magnetic vector potential $A$ analogously. These definitions automatically satisfy Equations \ref{eq:incompressibility} and \ref{eq:solenoid} for axisymmetric disturbances. We note that in the linearized equations, streamfunctions of the form $u_x = \partial_z \Psi$, $u_z = -(\partial_r + \frac{1}{r}) \Psi$, and the corresponding definitions of the magnetic vector potential, are convenient choices, but we make a different choice for this nonlinear investigation because of the incommutability of $\partial_r$ and $\partial_r + \frac{1}{r}$.

@@

Thus through judicious choice of cylinder rotation rates, one can set $q(r_0) = 3/2$, for quasi-Keplerian flow. Note that the narrow gap approximation imposes a linear shear (constant $q$), and thus the interaction of fluid perturbations with the base velocity profile differs significantly from the case considered here. Our base velocity is

\beq
\label{eq:baseu0}
\mathbf{u_0} = r\Omega(r) \phihat.
\eeq

We initialize a magnetic field

\beq
\label{eq:baseB0}
\mathbf{B_0} = B_0 \zhat + B_0 \xi \frac{r_0}{r} \phihat,
\eeq

so that the base magnetic field is axial when $\xi = 0$ and otherwise helical.

In this work we will focus our findings on two fiducial parameter sets, one for the standard MRI (hereafter SMRI) where $\xi = 0$ and one for the helical MRI (hereafter HMRI). We choose the SMRI parameters to be comparable to the case considered in \citei{Goodman:2002ix}. The HMRI parameters were chosen to be comparable to \citei{Hollerbach:2005tr}. Our fiducial parameters are described in Table \ref{table:parameters}.

\begin{widetext}
Our perturbed system is

@@

and $B = 0$ at $r = r_1, r_2$ (see \citei{Willis:2002bh}). Here, $I_n$ and $K_n$ are the modified Bessel functions of the first and second kind, respectively.

@@

We find the marginal system as a function of the dimensionless parameters. Marginality for our standard MRI system is a hyperplane in $(\reym, Pm, \beta, \Omega_2/\Omega_1, \mathrm{R}_1/\mathrm{R}_2$), but we hold all of these constant except for $\reym$. To analyze the MRI system at marginality, we fix the parameters listed in Table \ref{table:parameters} and determine the critical $\reym$ and $k_z$ by repeatedly solving the linear MRI system to determine the smallest parameter values for which the fastest growing mode is zero. That is, we solve the linear eigenvalue problem for eigenvalues $\sigma = \gamma + i \omega$. Figure \ref{fig:growth_rates} shows linear MRI growth rates $\gamma$ in the $(\reym, k_z)$ plane. For the fiducial standard MRI parameters in Table \ref{table:parameters} we find critical parameters $\reym_c = 3.30$ and $k_c = 0.901$.

Just as in the weakly nonlinear analyses of \citei{Umurhan:2007hs} and \citei{Clark:2016}, we tune the system away from marginality by taking $B_0 \rightarrow B_0\left(1 - \epsilon^2\right)$, where $\epsilon \ll 1$. We parameterize scale separation as $Z = \epsilon z$ and $T = \epsilon^2 t$, where $Z$ and $T$ are slowly varying spatial and temporal scales, respectively. We group the fluid variables into a state vector $\mathbf{V} = \left[\Psi, u, A, B\right]^{\mathrm{T}}$, such that the full nonlinear system in Equations \ref{eq:Psi_perturbed} - \ref{eq:nonlinear_B} can be expressed as

@@

where $\mathcal{D}$, $\mathcal{L}$, and $\widetilde{\mathcal{G}}$ are matrices defined in Appendix \ref{app:basic_equations}, and $\mathbf{N}$ is a vector containing all nonlinear terms defined in Appendix \ref{app:nonlinear_terms}. We expand the variables in a perturbation series $\mathbf{V} = \epsilon \mathbf{V}_1 + \epsilon^2 \mathbf{V}_2 + \epsilon^3 \mathbf{V}_3 + h.o.t.$ The perturbed system can then be expressed at each order by the equations

@@

See Appendix \ref{app:basic_equations} for the definition of matrices and nonlinear vectors, and a thorough derivation. We emphasize that Equations \ref{eq:ordere} - \ref{eq:ordere3} have the same form as these equations in the narrow gap case, although the matrices, which contain all radial derivatives, are significantly different in this wide gap formulation. This is because we do not have slow variation in the radial dimension. The slow variation in $Z$ and $T$ are parameterized as an amplitude function $\alpha(Z, T)$ which modulates the flow in these dimensions. This parameterization coupled with the boundary conditions lead us to an ansatz linear solution $\mathbf{V}_1 = \alpha(Z, T) \mathbb{V}_{11}(r) e^{i k_z z} + c.c.$, where the radial variation is contained in $\mathbb{V}_{11}$.

We solve the equations at each order using Dedalus, an open source pseudospectral code (Burns et al. in prep). We solve the radial portion of the eigenvectors on a basis of Chebyshev polynomials subject to our radial boundary conditions. %Because our domain is bounded only radially and not vertically, we solve the radial dimension on a basis of Chebyshev polynomials, and the vertical dimension on a Fourier series basis. 
We solve Equation \ref{eq:ordere} as a linear eigenvalue problem, and Equation \ref{eq:ordere2} as a linear boundary value problem. %should i mention spurious eigenvals? 
%To analyze the MRI system at marginality, we fix the parameters listed in Table \ref{table:parameters} and determine the critical $\reym$ and $k_z$ by repeatedly solving Equation \ref{eq:ordere} to determine the smallest parameter values for which the fastest growing mode is zero. That is, we solve the linear eigenvalue problem for eigenvalues $\sigma = \gamma + i \omega$. Figure \ref{fig:growth_rates} shows linear MRI growth rates $\gamma$ in the $(\reym, k_z)$ plane. For the fiducial standard MRI parameters in Table \ref{table:parameters} we find critical parameters $\reym_c = 0.84$, $k_c = 0.0127$.

@@

We similarly define appropriate boundary conditions for the insulating case. We take the inner product of the adjoint homogenous solution with the terms in Equation \ref{eq:ordere3} that are resonant with $e^{i k_c z}$. This gives us

\beq
\langle \mathbb{V}^\dagger | \mathcal{D} \mathbb{V}_{11} \rangle \partial_T \alpha + \langle  \mathbb{V}^\dagger | \widetilde{\mathcal{G}} \mathbb{V}_{11} - \xi \widetilde{\mathcal{H}} \mathbb{V}_{11} \rangle \alpha +  \langle \mathbb{V}^\dagger | \widetilde{\mathcal{L}}_1 \mathbb{V}_{21} + \widetilde{\mathcal{L}}_2 \mathbb{V}_{11} + \xi \mathcal{H} \mathbb{V}_{21} \rangle \partial_Z^2 \alpha = \langle \mathbb{V}^\dagger | \mathbf{N}_{31} \rangle \alpha |\alpha|^2,
\eeq

or Equation \ref{eq:gle}, the Ginzburg-Landau Equation, where the coefficients are

\beq
b = \langle  \mathbb{V}^\dagger | \widetilde{\mathcal{G}} \mathbb{V}_{11} - \xi \widetilde{\mathcal{H}} \mathbb{V}_{11} \rangle / \langle \mathbb{V}^\dagger | \mathcal{D} \mathbb{V}_{11} \rangle,
\eeq

\beq
h = \langle \mathbb{V}^\dagger | \widetilde{\mathcal{L}}_1 \mathbb{V}_{21} + \widetilde{\mathcal{L}}_2 \mathbb{V}_{11} + \xi \mathcal{H} \mathbb{V}_{21} \rangle / \langle \mathbb{V}^\dagger | \mathcal{D} \mathbb{V}_{11} \rangle,
\eeq

\beq
c = \langle \mathbb{V}^\dagger | \mathbf{N}_{31} \rangle / \langle \mathbb{V}^\dagger | \mathcal{D} \mathbb{V}_{11} \rangle. 
\eeq
